---
title: "Experimental Design and Data Analysis: Assignment 2"
author: "Rick van Ravensberg, Manuel Giurea Scanferla & Ángel Gil, group 64"
date: "15 March 2023"
output: pdf_document
fontsize: 10pt
highlight: tango
editor_options: 
  markdown: 
    wrap: 72
---

```{r,echo=FALSE}
options(digits=3)
```

## Exercise 1. Trees

**a)** We begin by generating the ANOVA table without taking diameter
and height into account:

```{r,echo = FALSE}
data=read.table(file="Data/treeVolume.txt",header=TRUE)

beech <- subset(data, type == "beech")$volume
oak <- subset(data, type == "oak")$volume

result <- lm(volume~type,data=data)
anova(result)

```

The "Pr(\>F)" column shows the p-value associated with the F statistic.
In this case, the p-value is 0.17, which is greater than the
significance level of 0.05. Therefore, we fail to reject the null
hypothesis and conclude that tree type does not influence volume. A
t-test can be thought of as a special case of ANOVA, where there are
only two groups being compared. In this case, we have two groups (Beech
and Oak), so we could perform a two-sample t-test to compare their mean
volumes.

```{r,echo = FALSE}
t.test(beech, oak)
```

The null hypothesis is that the means of the two groups are equal, and
the alternative hypothesis is that they are not equal and since p-value
is 0.2, there is not enough evidence to reject the null hypothesis at a
significance level of 0.05. This confirms the results of the ANOVA test.
We can then estimate the volumes of the two types of trees by looking at
their mean:

```{r}
mean(beech) # mean volume of type beech
mean(oak) # mean volume of type oak
```

Lastly, we want to check whether the assumption of normality is not
totally untrue using a qqnorm plot and plotting the fitted values
against the residuals:

```{r, fig.margin =  c(0, 0, -5, -5),fig.width=8,fig.height=4, fig.align="center", echo=FALSE}
par(mfrow=c(1,2)); qqnorm(residuals(result),main = NULL)
plot(fitted(result),residuals(result))
```

we can notice that QQ-norm does not follow a straight line, hence we can
not confirm the normality assumption.

**b)** To investigate whether the influence of diameter and height on
volume is similar for both tree types (Beech and Oak), we can fit two
linear regression models.

```{r}
#pairwise interaction between type and diameter
mod1 <- lm(volume~height+type*diameter, data=data)
anova(mod1)
```

```{r}
#pairwise interaction between type and height
mod2 <- lm(volume~diameter+type*height, data=data)
anova(mod2)
```

Looking at the p-values, from both models we do not reject the null
hypothesis and conclude that the effect of diameter and height on volume
is similar for both tree types. We want now to try and visualize the
relationship between volume and diameter/height for each tree type:

```{r, fig.margin = TRUE,fig.width=8,fig.height=4,fig.align="center", echo=FALSE}

par(mfrow=c(1,2))

library(ggplot2)
ggplot(data, aes(x = diameter, y = volume, color = type)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle("Volume vs Diameter") +
  labs(color = "Tree Type")

ggplot(data, aes(x = height, y = volume, color = type)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle("Volume vs Height") +
  labs(color = "Tree Type")

```

From the plots, it seems like there may be a different relationship
between volume and diameter/height for Beech and Oak trees. It appears
to exist a strong relationship between volume and diameter for both
trees. Similarly, also between volume and height we can see a
significant relationship, in this case mostly for beech trees. Let's now
fit two more linear regression models and compare their coefficients to
confirm these observations:

```{r}
# Linear regression model for volume vs diameter for each tree type
beechModel <- lm(volume ~ diameter, data = data, subset = type == "beech")
oakModel <- lm(volume ~ diameter, data = data, subset = type == "oak")

# Compare the coefficients
summary(beechModel)$coefficients
summary(oakModel)$coefficients

# Linear regression model for volume vs height for each tree type
beechModel2 <- lm(volume ~ height, data = data, subset = type == "beech")
oakModel2 <- lm(volume ~ height, data = data, subset = type == "oak")

# Compare the coefficients
summary(beechModel2)$coefficients
summary(oakModel2)$coefficients
```

**c)** Based on the linear regression models, we can see that both
diameter and height have a positive influence on volume for both tree
types. We already saw that type does not influence the volume. The
coefficients for diameter are similar for both tree types, with beech
having a slightly higher coefficient (5.07) compared to oak (4.74). This
suggests that diameter has a similar influence on volume for both beech
and oak trees. The coefficients for height are also positive for both
tree types, but the magnitudes are different. The beech trees'
coefficient (1.54) is much higher than that for oak trees (0.31),
indicating that height has a stronger influence on volume for beech
trees compared to oak trees. We can also notice it from the previous
graph. Overall, we see that increasing both height and diameter result
in an increase in volume, especially diameter for both type of trees. To
predict the volume for a tree with the overall average diameter and
height, we need to calculate the mean values for these variables:

```{r,echo = FALSE}
mean_diameter <- mean(data$diameter)
mean_height <- mean(data$height)

# Predict volume for beech tree
mean_volume_beech <- predict(beechModel, data.frame(diameter = mean_diameter, height = mean_height))

# Predict volume for oak tree
mean_volume_oak <- predict(oakModel, data.frame(diameter = mean_diameter, height = mean_height))

mean_volume_beech
mean_volume_oak
```

**d)** Since we know that the formula to calculate the volume of a
cylinder is $v=\frac{(\pi*d^2*h)}{4}$, we simplify and substitute
diameter with $diameter^2$ in the model. We also remove type in the
model, since we already understood that is not significant.

```{r}
# previous model
mod3=lm(volume~diameter+height,data=data)
summary(mod3)

# new model
data$diam_sq=data$diameter^2
mod4=lm(volume~diam_sq+height,data=data)
summary(mod4)
```

We can confirm that the model with the transformed explanatory variable
is better. This is indicated by the higher F-value and lower p-value for
the diam_sq variable compared to the diameter variable in the first
model. Additionally, the multiple R-squared and adjusted R-squared
values are higher in the second model, indicating a better fit. Lastly,
we want to check the assumption of normality to understand if we can
make valid assumptions from this model:

```{r, fig.margin = c(0, 0, -5, -5),fig.width=8,fig.height=4,fig.align="center", echo=FALSE}
par(mfrow=c(1,2));par(cex = 0.7); qqnorm(residuals(mod4),main = NULL)
plot(fitted(mod4),residuals(mod4))

```

We can see that the norm qq-plot follows a straight line and the
fitted-residual model does not show a pattern. This means that the
normality is checked and we can draw valid conclusions from the model.

## Exercise 2. Expenditure on criminal activities

```{r, include=FALSE}
# Read data
crimedf <- as.data.frame(read.table("data/expensescrime.txt", header=TRUE))
head(crimedf)
response <- "expend"
exp_vars <- c("bad", "crime", "lawyers", "employ", "pop")
```

**a)** In order to check the pairwise interactions between all the
variables of interest, we can use scatter plots:

```{r, echo=FALSE, fig.height=5, fig.width=5, fig.align="center"}
par(cex = 0.8);pairs(crimedf[, c(response, exp_vars)])
```

After examining the pairs plot, it is evident that there are linear
relationships among the predictors, namely *bad*, *lawyer*, *employ*,
and *pop*, which can lead to the issue of collinearity. Moreover, a
strong positive correlation is observed between the response variable
expend and the aforementioned predictors, except for *crime*. However,
there are some outliers present in the higher range of the data, which
can potentially impact the model's accuracy. The existence of
collinearity among the predictors is a concern as it can make it
challenging to estimate the regression coefficients due to redundant
information.

In order to find influence points, we can use Cook's distance, where a
distance greater than 1 indicates an outlier:

```{r}
crimelm <- lm(expend~bad+crime+lawyers+employ+pop, data=crimedf)
```

```{r, fig.margin = TRUE,fig.width=6,fig.height=3,fig.align="center", echo=FALSE}
par(cex = 0.8);plot(cooks.distance(crimelm), type='b', ylab="Distance", main = "Cook's distance for expensecrime")
```

```{r}
cooks.distance(crimelm)[cooks.distance(crimelm) > 1]
```

Indices 5, 8, 35 and 44 are thus regarded as influence points and can be
removed from the data:

```{r}
Updcrimedf <- crimedf[-c(5,8,35,44),]
```

Let's now investigate collinearity by examining the correlations between
all the explanatory variables:

```{r, echo=FALSE}
knitr::kable(round(cor(crimedf[, c(exp_vars)]), 2), caption = "Correlation coefficients")
```

The correlation table confirms what was observed in the pairs plot;
there are strong correlations between the explanatory variables,
excluding crime. The lowest correlation coefficient is 0.85, between bad
and lawyers, indicating that all predictors (excluding crime) contain
comparable information.

We can also use the VIF to see which variables are collinear, where a
value \>5 is cause for concern:

```{r, echo=FALSE}
library(car)
```

```{r}
vif(lm(expend~bad+crime+lawyers+employ+pop, data=crimedf))
```

This result also supports that the variables *bad*, *lawyers*, *employ*
and *pop* are collinear.

**b)** From now on, we will proceed without the influence points.

```{r, include=FALSE}
library(dplyr)
```


```{r}
# level 1
mbad <- lm(expend~bad, data=Updcrimedf); mcrime <- lm(expend~crime, data=Updcrimedf)
mlawyers <- lm(expend~lawyers, data=Updcrimedf); memploy <- lm(expend~employ, data=Updcrimedf)
mpop <- lm(expend~pop, data=Updcrimedf)
l1m <- list(mbad, mcrime, memploy, mlawyers, mpop); 
var1 <- c("bad", "crime", "employ", "lawyears", "pop")
Rsquared1 <- c()
for(m1 in l1m){
  r_squared <-  summary(m1)$r.squared
  Rsquared1 <- c(Rsquared1, r_squared)
}
knitr::kable(data.frame(Predictor = var1, Rsquared = Rsquared1), 
             caption = "Step-up: Level 1")
```

From Table 2, it is deduced that *employ* is the variable to be included
in the model. Analogously, we obtain Tables 3 and 4 which allow us to
obtain the best step-up model.

```{r, echo=FALSE}

# level 2
mbad2 <- lm(expend~employ+bad, data=Updcrimedf); 
mcrime2 <- lm(expend~employ+crime, data=Updcrimedf)
mpop2 <- lm(expend~employ+pop, data=Updcrimedf); 
mlawyers2 <- lm(expend~employ+lawyers, data=Updcrimedf)
lv2_models <- list(mbad2, mcrime2, mlawyers2, mpop2); 
variables_2 <- c("employ+bad", "employ+crime", "employ+lawyears", "employ+pop")
rsq2 <- c()
significance_2 <- c()
for(model_1 in lv2_models){
  is_significant <- sum(c(summary(model_1)$coefficients[,4] < 0.05)) > 2 
  r_squared <-  summary(model_1)$r.squared
  rsq2 <- c(rsq2, r_squared)
  significance_2 <- c(significance_2, is_significant) 
}
# level 3
mbad3 <- lm(expend~employ+crime+bad, data=Updcrimedf);
mpop3 <- lm(expend~employ+crime+pop, data=Updcrimedf);
mlawyers3 <- lm(expend~employ+crime+lawyers, data=Updcrimedf);
lv3_models <- list(mbad3, mlawyers3, mpop3); 
variables_3 <- c("employ+crime+bad", "employ+crime+lawyers", "employ+crime+pop")
rsq3 <- c()
significance_3 <- c()
for(model_1 in lv3_models){
  is_significant <- sum(c(summary(model_1)$coefficients[,4] < 0.05)) > 3 
  r_squared <-  summary(model_1)$r.squared
  rsq3 <- c(rsq3, r_squared)
  significance_3 <- c(significance_3, is_significant) 
}
par(mfrow=c(1,2)); knitr::kable(data.frame(Predictor = variables_2, r.squared = rsq2, 
  Significant = significance_2), caption = "Step-up: Level 2")
knitr::kable(data.frame(Predictor = variables_3, r.squared = rsq3, 
  Significant = significance_3), caption = "Step-up: Level 3")
```

The variables added in order were *employ*, *crime* and *pop*, after
which no further added variables had significant p-values (notice that
we omitted the last level). Hence the final model is as follows:

```{r}
mstepup <- lm(expend~employ+crime+pop, data=Updcrimedf)
summary(mstepup)
```

Step-up final model: expend = -247 + 0.0209\*employ + 0.0543\*crime +
0.0714\*pop $\pm$ error, with $R^2 = 0.974$. We can check VIF for
further insight:

```{r}
vif(mstepup)
```

Although the step-up method naturally removes collinearity and produced
a better model than the one obtained in *a)*, which had an R-squared
value of 0.957, there is still a high VIF value for *employ* and *pop*.

Finally, we check the model assumptions:

```{r, fig.margin = c(0, 0, -5, -5),fig.width=7,fig.height=3.5,fig.align="center", echo=FALSE}
par(mfrow=c(1,2));par(cex = 0.7);
qqnorm(residuals(mstepup),main=NULL)
plot(fitted(mstepup), residuals(mstepup))
```

We can assume normality, although it is worth addressing that the qq
plot makes small curved groups of points.

**c)** Using the step-up model from the previous case, the 95%
prediction interval for *expend* is given by:

```{r}
newdata <- data.frame(bad=50, crime=5000, lawyers=5000, employ=5000, pop=5000)
predict(mstepup, newdata, interval="prediction")
```

One way to improve this interval would be by transforming the existing
variables to better capture their relationship with the response
variable. However, this would require additional analysis and
experimentation. Moreover, we already removed the outliers and the
inclusion of more variables in the model is not recommended, since we
already tested that there are no significant variables to add.

**d)** Let's apply the LASSO method:

```{r, include=FALSE}
# install.packages("glmnet", repos = "https://cran.us.r-project.org")
library(glmnet)
```

```{r}
x <- as.matrix(Updcrimedf[, exp_vars])
y <- as.matrix(Updcrimedf[, c(response)])
train <- (sample(1:nrow(x), 0.67*nrow(x))) # train by using 2/3 of the data
x.train <- x[train,]; y.train <- y[train]  # data to train
x.test <- x[-train,]; y.test <- y[-train]  # data to test the prediction quality
mlasso <- glmnet(x.train, y.train, alpha=1)
cvlasso <- cv.glmnet(x.train,y.train,alpha=1,type.measure='mse')
```

```{r,fig.width=6,fig.height=3,fig.align="center"}
plot(mlasso, label=T, xvar="lambda")  # visualization of the lasso path
```

```{r,fig.width=6,fig.height=3,fig.align="center"}
plot(cvlasso) # the best lambda by cross-validation
```

```{r}
(lambda.1se <- cvlasso$lambda.1se) #  lambda for most regularized model
```

Let's now asses the performance of the model built using Lasso
regression.

```{r}
assess.glmnet(mlasso, newx = x.test, newy = y.test, s=cvlasso$lambda.1se)
```

In order to compare this model to the step-up model from (b), we find
the MSE for the latter:

```{r}
newdata <- data.frame(x.test)
y.pred <- predict(mstepup, newdata, interval="confidence", level=0.95)
msestepup <- mean((y.test - y.pred)^2); msestepup
```

After several runs (of the Lasso method), we can conclude that the
step-up model outperforms it consistently in this case (20 out of 20
tries, but we did not keep track), producing a smaller MSE. This could
happen because LASSO is better suited to situations with many more
explanatory variables.

## Exercise 3. Titanic

**a)** To get a better understanding we visualized several elements from
the data. First, we were interested in the number of people per class.
It appears that the third class is relatively large and the first and
second class are about equal size. Then we visualized the effect of age
and sex on survival rate. The figure shows that the average age of
surviving men is lower than the average age of the men that did not
survive, this indicates a positive effect of age survival rate. However,
females show the opposite effect, disputing the previous indication.
Some of the data points missed the age column, these values were
replaced with the average age.

```{r, echo=FALSE,fig.width=6,fig.height=3,fig.align="center"}
titanicdata = read.table(file="Data/titanic.txt", header=TRUE)

pclassfreq <- table(titanicdata$PClass)
barplot(pclassfreq, main = "Passengers per Class")
```

```{r, echo=FALSE,fig.width=6,fig.height=4}
par(cex = 0.8);boxplot(Age ~ Survived + Sex,
        data = titanicdata,
        col = rainbow(2, s = 0.5),
        notch = TRUE,
        main = "Age Distribution",
        names = c("F - D", "F - S", "M - D", "M - S"),
        xaxs = FALSE
)
legend("topleft", fill = rainbow(2, s = 0.5), legend = c("Diseased", "Survived"))
```

A logistic regression model was created to investigate the association
between the survival status and the predictors *PClass*, *Age* and
*Sex*, according to code in Listing 6. The estimated odds is calculated
as follows:

$$
\hat{o}_k = \frac{Pr(Y_k = 1)}{Pr(Y_k = 0)} = exp\{ 3.76 -1.29 Class2_k -2.52 Class3_k -0.039177 Age_k -2.63 SexMale_k \}
$$

First of all, we see that age does not strongly effect the survival
rate. Next, we observe a relatively high negative value for being a
third class passenger, this indicates that traveling in third class
negatively effects survival rate. The parameter for second class
passengers also shows a negative influence on the survival rate, but the
effect is not as strong as for the third class passengers. Finally, we
see that being a male also has a strong negative effect on the
probability of survival.

```{r}
titanicmodel = glm(Survived ~ PClass + Age + Sex, data=titanicdata, family=binomial)
summary(titanicmodel)
```

**b)** The following different models were used to investigate the
interaction between different predictors. All models use a null
hypothesis $H_0 : \beta_{Age} = \beta_{Class}$ and a significance level $\alpha = 0.05$.

```{r}
modelAgePClass = glm(Survived ~ Age * PClass, data=titanicdata, family=binomial)
anova(modelAgePClass, test="Chisq")
    
modelAgeSex = glm(Survived ~ Age * Sex, data=titanicdata, family=binomial)
anova(modelAgeSex, test="Chisq")
```

The first model, which models the interaction between *Age* and *Class*,
returns a p-value of 0.558. This does not reject *H*~0~, meaning that
there is no interaction between predictor *Age* and factor *Class*. The
second model on the interaction between *Age* and *Sex* returns p \<
0.001. This value rejects our *H*~0~, which means that the two predictors
interact.

From this point, we work with a model that uses the *PClass* predictor,
but it also integrates the fact that *Age* and *Sex* interact.

```{r}
modelFinal = glm(Survived ~ PClass + Age * PClass, data=titanicdata, family=binomial)
anova(modelFinal, test="Chisq")
```

The estimate of the probability of survival for each combination of
levels of the factors *Class* and *Sex* for a person of age 53 can be
found in the following table.

|         | Male      | Female    |
|---------|-----------|-----------|
| Class 1 | 0.5358867 | 0.7119470 |
| Class 2 | 0.5115599 | 0.6754746 |
| Class 3 | 0.5035067 | 0.6097319 |

: Probability of survival for each combination of levels of the factors
Class and Sex for a person of age 53

```{r}
titanicdata$PClass = factor(titanicdata$PClass)
titanicdata$Sex = factor(titanicdata$Sex)

summary(modelFinal)

m_names = list(c("Class1", "Class2","Class3"),c("Male", "Female"))
results = matrix(, nrow=3, ncol=2, dimnames=m_names)

classLabels = c("1st", "2nd", "3rd")
genderLabels = c("male", "female")
for (i in seq_along(classLabels)) {
  for (j in seq_along(genderLabels)) {
    newData = data.frame(Age=53, Sex=genderLabels[j], PClass=classLabels[i])
    x = predict(modelFinal, newData, type="response")
    results[i, j] = 1/(1 + exp(-x))
  }
}
```

**c)** The second approach is not wrong, contingency tables should
produce the similar conclusions as the logistic regression model. The
logistic regression model is a rather complex model, it does not only
test for predictor interaction, but it also creates a model which can be
used for prediction. So the logistic regression model is more versatile
than the contingency table. The complexity of the model can also be
considered as a disadvantage, as it is less intuitive to interpret.

We propose to split the data into a training and a test set to assess
the performance of our predictor. The algorithm can use the training set
to train the model, i.e. it tries to find the optimal parameters. The
test set contains data points which the model has never seen before,
these data points are used to assess the performance of the model. If
the model is able to accurately predict the survival on unseen data,
then we can conclude that the model generalizes the data well.

**d)** We created contingency tables to investigate whether factor
passenger class has an effect on the survival status, see Table 6. The
built-in function `chisq.test(classTable)` shows a result of p \< 0.001,
indicating that passenger class and survival are dependent.

|          | Class 1 | Class 2 | Class 3 |
|----------|---------|---------|---------|
| Survived | 193     | 119     | 138     |
| Deceased | 129     | 161     | 573     |

: Number of passengers per class against survival

A similar thing was done for the factor gender, see Table 7. Because
this is a 2x2 table, we can use the Fisher's exact test. This is a
built-in function in R, `fisher.test(genderTable)`, it also reports a
result of p \< 0.001, thus the gender and survival rate are also
dependent.

|          | Male | Female |
|----------|------|--------|
| Survived | 142  | 308    |
| Deceased | 709  | 154    |

: Number of passengers per gender against survival

**e)** The second approach is not wrong, contingency tables should
produce the similar conclusions as the logistic regression model. The
logistic regression model is a rather complex model, it does not only
test for predictor interaction, but it also creates a model which can be
used for prediction. So the logistic regression model is more versatile
than the contingency table. The complexity of the model can also be
considered as a disadvantage, as it is less intuitive to interpret.

## Exercise 4. Military coups

**a)** All variables of the coups data set were loaded in a Poisson GLM,
using the following procedure.

```{r}
coupsdata=read.table(file="Data/coups.txt", header=TRUE)
coupsglm=glm(miltcoup ~ oligarchy + pollib + parties +
               pctvote + popn + size +numelec + numregim,
    family=poisson,
    data=coupsdata )
summary(coupsglm)
```

These results indicate that the military coups are significantly
predicted by 3 factors: *oligarchy*, *pollib* and *parties*.

1.  The number of years the country is ruled by a military oligarchy has
    a positive effect on the number of military coups; an increase in
    years of military oligarchy predicts an increase in military coups.

2.  The political liberalization level has a negative effect on the
    number of military coups; the fewer civil rights a country has, the
    number of coups increases.

3.  The number of political parties has a positive effect on the number
    of military coups; as such that when a country has more political
    parties, a greater number of military coups can be expected.

All other variables in the model didn't have a significant effect on the
number of military coups.

**b)** The number of exploratory variables in the former Poisson
regression was reduced by removing the variable that has the highest
P-value in the model. This procedure was repeated until all variables in
the model are significant. The following variables were successively
removed from the model: *numelec*, *numregim*, *size*, *popn*,
*pctvote*. The final model contains the following explanatory variables:
*oligarchy*, *pollib*, *parties*.

```{r}
coupsglm5 <- step(coupsglm, direction = "backward", trace = FALSE, family ="poisson")
summary(coupsglm5)
```

While in the full model *oligarchy* was barely significant with a value
of *p* = 0.034, in the reduced model this variable has a very small
p-value (*p* \< 0.001), which indicates that the presence of an oligarchy
is a valid predictor of a military coup. The predictive value of the
number of political parties in a country reduced in significance. The
political liberalization level remained similar in significance.

**c)** Using the model from the previous section, the number of coups
can be predict for each of the levels of political liberalization. For
the variables *oligarchy* and *parties* averages were calculated and fed
into the model, respectively 5.222 and 17.083.

```{r}
# Level 0 No civil rights for political expression
pollib0 = data.frame(
    pollib=0,
    oligarchy=mean(coupsdata[,'oligarchy']),
    parties=mean(coupsdata[,'parties'])
)
predict(coupsglm5, newdata=pollib0, type="response")

# Level 1 Limited civil rights for expression
pollib1 = data.frame(
    pollib=1,
    oligarchy=mean(coupsdata[,'oligarchy']),
    parties=mean(coupsdata[,'parties'])
)
predict(coupsglm5, newdata=pollib1, type="response")

# Level 2 Full civil rights
pollib2 = data.frame(
    pollib=2,
    oligarchy=mean(coupsdata[,'oligarchy']),
    parties=mean(coupsdata[,'parties'])
)
predict(coupsglm5, newdata=pollib2, type="response")
```

From these predictions can be concluded that more military coups are
expected if less civil rights are granted in the average African
country. When no civil rights exist, the predicted number of coups is a
little over 3, while with full civil rights, the predicted number is
just a little under 1.